Listings of equations, figures, tables and sections of the article '{\it \bf Musical elements in the discrete-time representation of sound}' and of the scripts in the MASS toolbox

### 2017-08-05 article/acm/body.tex
enhanced disclaimer, equations and notation

The representation of sound can take many forms,
from musical scores and texts in a phonetic language to electric analog signals and binary data.
It includes arbitrary sets of features and spectral decompositions such as obtained through wavelets.
Although the terms 'audio' and 'sound' are often used without distinction
and 'audio' has many definitions which depend on the context and the author,
audio most often means a representation of sound by means
of the wave amplitude through time.
In this sense,
audio expresses sonic waves from direct synthesis or input by microphones,
although these sources are not always neatly distinguishable
e.g. as captured sounds are processed to generate new sonorities.
Digital audio protocols often implies quality loss (to yield smaller files, ease storage and transfer)
and are called \emph{lossy}~\cite{loss}.
This is the case e.g. of MP3 and Ogg Vorbis.
Non-lossy representations of digital audio, called \emph{lossless} protocols or formats,
on the other hand, assures perfect reconstruction of the analog wave within any convenient precision.
The standard paradigm of lossless audio consists of
representing the sound with samples equally spaced
by a duration $\lambda_s$, and specifying the amplitude of each sample by a fixed number of bits.
This is the Pulse Code Modulation (PCM) representation of sound.
A PCM audio format has two essential attributes:
a sampling frequency $f_s=\frac{1}{\lambda_s}$ (also called e.g. sampling rate or sample rate),
which is the number of samples used for representing a second of sound;
and a bit depth, which is the number of bits for each sample.
Figure~\ref{fig:PCM} shows $25$ samples of a PCM audio with a bit depth of $4$,
which yields $2^4=16$ possible values for the amplitude of each sample
and a total of $4 \times 25= 100$ bits for representing the whole sound.

@@

The objectives of this paper are:

\begin{enumerate}
	\item Present a concise set of mathematical relations between musical basic elements and sequences of PCM audio samples.
	\item Introduce a framework of sound and musical synthesis with control at sample level which entails potential uses in psychoacoustic experiments and synthesis with extreme precision (recap in Section~\ref{cap:conclusao}).
	\item Provide a powerful theoretical framework which can be used to synthesize musical pieces and albums.
	\item Provide approachability to the developed framework~\footnote{All
		the analytic relations presented in this article are implemented as public domain scripts,
		i.e. small computer programs using accessible technologies for distribution and validation.
		This constitutes the \massa\ toolbox, available in an open Git repository~\cite{gitBook}.
		These scripts are written in Python and make use of Numpy,
		which performs numerical routines efficiently (e.g. through LAPACK).
		Part of the scripts have been transcribed to JavaScript and native Python,
		which favors their use in Web browsers such as Firefox and
		Chromium~\cite{numpy, audiolab, tutpython, python}.
		These are all open technologies, published using licenses that allow for copy,
		distribution and use for research and generating derivatives.
		Hence, the work presented here aims at being compliant with recommended
		practices for availability and validation
		and should ease co-authorship processes~\cite{Raymond,Lessig}.}.
	\item Provide a didactic presentation of the content, which is highly multidisciplinary,
		involving signal processing, music, psycho-acoustics and programming.
\end{enumerate}

@@

\subsection{Synonymy and polysemy (disclaimer)}
Given that the main topic of this article (the expression of musical elements
in PCM audio) is multidisciplinary and involves art,
the reader should be aware that many of the vocabulary admit different choices of terms and definitions.
More specifically, it is often often the case where many words can express the same concept
and where one word can carry different meanings.
This is a very deep issue which might receive a dedicated manuscript.
The reader might need to read the rest of this document to understand this
small selection of synonymy and polysemy in the literature,
but it is important to illustrate the point before the more dense sections:
\begin{itemize}
	\item A ``note'' can mean a pitch or an abstract construct with pitch and duration or a sound emitted from a musical instrument or a specific note in a score or a music.
	\item The sampling rate (discussed above) is also called the sampling frequency.
	\item An harmonic in a sound is most often a sinusoidal component which is is the harmonic series of the fundamental frequency. Many times, however, the terms harmonic and component are not distinguished. A harmonic can also be a note performed in a instrument by preventing certain overtones (components).
	\item Harmony can refer to chords or to note sets related to chords or even to ``harmony'' in a more general sense, as a kind of balance and consistency.
	\item A ``tremolo'' can mean different things, e.g. in a piano score, a tremolo is a fast alternation of two notes (pitches) while in computer music theory it is (most often) an oscillation of intensity.
\end{itemize}

We strived to avoid nomenclature clashes and the use of more terms and needed.
Also, some of these issues are exposed when the context is convenient, such as in the discussion of timbre.

@@

\subsection{Duration}
The sample frequency $f_s$ is defined as the number of samples in each second of the discrete-time signal. Let $T_i=\{t_i\}$ be an ordered set of real samples separated by $\delta_s=1/f_s$ seconds.
A musical note of duration $\Delta$ seconds can be expressed as a sequence $T_i^{\Delta}$ of $\lfloor \Delta . f_s \rfloor$ samples.
That is, the integer part of the multiplication is considered, and an error of at most $\delta_s$ missing seconds is admitted, which is usually fine for musical purposes. As an example, if $f_s=44.1kHz \;\;\Rightarrow\;\;\lambda_s=\frac{1}{44100}\approx 23$ microseconds. Thus:

@@

\noindent where $\lambda_f=\frac{f_s}{f}=\frac{\delta_f}{\lambda_s}$ is the number of samples in the period.

@@

Let $e_i$ be integers indexed by $i$ such that
$f=\varepsilon^{e_i} f_0$,
where $f_0$ is any fixed frequency.
The symmetric scales mentioned above can be expressed as:
\begin{equation}\label{escSim}
\begin{split}
\text{chromatic}    & = E_i^c = \{e_i^c\}_0^{11} = \\
                    & =  \{0,1,2,3,4,5,6,7,8,9,10,11\} = \{i\}_0^{11}\\
	\text{whole tones}  & = E_i^{wt} = \{e_i^{wt}\}_0^{5} = \{0,2,4,6,8,10\} = \{2.i\}_0^{5} \\
\text{minor thirds} & = E_i^{mt} = \{e_i^{mt}\}_0^{3} = \{0,3,6,9\} = \{3.i\}_0^3 \\
\text{major thirds} & = E_i^{Mt} = \{e_i^{Mt}\}_0^{2} = \{0,4,8\} = \{4.i\}_0^2\\
\text{tritones}     & = E_i^{tt} = \{e_i^{tt}\}_0^{1} = \{ 0, 6 \} = \{6.i\}_0^1
\end{split}
\end{equation}

@@

\begin{equation}\label{eq:escalas}
\begin{split}
\text{aeolian}    & = \text{natural minor scale} = \\
                  & = E_i^m = \{e_i^m\}_0^6 = \{0,2,3,5,7,8,10\} \\
\text{locrian}    & = E_i^{lo} = \{e_i^{lo}\}_0^6 = \{0,1,3,5,6,8,10\} \\ 
\text{ionian}     & = \text{major scale} =  \\
                       & = E_i^M = \{e_i^M\}_0^6 = \{0,2,4,5,7,9,11\} \\
\text{dorian}     & = E_i^{d} = \{e_i^{d}\}_0^6 = \{0,2,3,5,7,9,10\} \\
\text{phrygian}   & = E_i^{p} = \{e_i^{p}\}_0^6 = \{0,1,3,5,7,8,10\} \\
\text{lydian}     & = E_i^{l}=\{e_i^{l}\}_0^6 = \{0,2,4,6,7,9,11\} \\
\text{mixolydian} & = E_i^{mi} = \{e_i^{mi}\}_0^6 = \{0,2,4,5,7,9,10\}
\end{split}
\end{equation}

@@

Although it is not a traditional scale, the harmonic series is often used as such::
\begin{equation}\label{eq:serieHarmonica}
\begin{split}
H_i = & \{h_i\}_0^{19}= \\
      & \{ 0,12,19+0.02,  24,28-0.14, 31+0.2, 34-0.31, \\
                     & 36, 38+0.04,40-0.14, 42-0.49, 43+0.02, \\
                     & 44+0.41, 46-0.31, 47-0.12, \\
                     & 48, 49+0.05, 50+0.04, 51-0.02, 52-0.14   \}
\end{split}
\end{equation}

@@

\begin{equation}\label{eq:groups}
\begin{split}
\forall \;\; p_1,p_2 \in G \Rightarrow  \quad   & p_1 \bullet p_2  = p_3 \in G \\ 
     & \text{(closure property)} \\
\forall \;\; p_1,p_2,p_3 \in G \Rightarrow \quad & (p_1\bullet p_2)\bullet p_3  = p_1\bullet (p_2\bullet p_3) \\
     & \text{(associativity property)} \\
\exists \;\; e \in G :                  \quad    & p \bullet e  = e \bullet p \;\;\;\; \forall p \in G  \\ 
     &  \text{(identity element existence)} \\
\forall \;\; p \in G, \;\exists\; p^{-1} :\quad  &  p\bullet p^{-1}i =p^{-1}\bullet p = e \\
     &  \text{(inverse element existence)}
\end{split}
\end{equation}

@@

The possibilities provided by this exposition pour from both the organization of knowledge and the ability to achieve sounds which are extremely true to the models.
For example, one can produce noises with arbitrary spectrum resolution and a musical note can be synthesized with the parameters (e.g. of a vibrato) updated sample-by-sample.
These abilities can be crucial e.g. for psychoacoustic experiments or for the development of
interfaces to generate high quality musical sounds or pieces.
Simply put, it is compelling
for many scientific, artistic and didactic purposes.
A fact evidenced by a number of uses this framework had in courses, in software implementations and in making music~\cite{vimeoLM,vivace,repoDissertacao,dissertacao}.

>>>>

The representation of sound can take many forms,
from musical scores and texts in a phonetic language to electric analog signals and binary data.
It includes arbitrary sets of features and spectral decompositions such as obtained through wavelets.
Although the terms 'audio' and 'sound' are often used without distinction
and 'audio' has many definitions which depend on the context and the author,
audio most often means a representation of sound by means
of the wave amplitude through time.
In this sense,
audio expresses sonic waves from direct synthesis or input by microphones,
although these sources are not always neatly distinguishable
e.g. as captured sounds are processed to generate new sonorities.
Digital audio protocols often implies quality loss (to yield smaller files, ease storage and transfer)
and are called \emph{lossy}~\cite{loss}.
This is the case e.g. of MP3 and Ogg Vorbis.
Non-lossy representations of digital audio, called \emph{lossless} protocols or formats,
on the other hand, assures perfect reconstruction of the analog wave within any convenient precision.
The standard paradigm of lossless audio consists of
representing the sound with samples equally spaced
by a duration $\delta_s$, and specifying the amplitude of each sample by a fixed number of bits.
This is the Pulse Code Modulation (PCM) representation of sound.
A PCM audio format has two essential attributes:
a sampling frequency $f_s=\frac{1}{\delta_s}$ (also called e.g. sampling rate or sample rate),
which is the number of samples used for representing a second of sound;
and a bit depth, which is the number of bits for each sample.
Figure~\ref{fig:PCM} shows $25$ samples of a PCM audio with a bit depth of $4$,
which yields $2^4=16$ possible values for the amplitude of each sample
and a total of $4 \times 25= 100$ bits for representing the whole sound.

@@

The objectives of this paper are:
\begin{enumerate}
	\item Present a concise set of mathematical relations between musical basic elements and sequences of PCM audio samples.
	\item Introduce a framework of sound and musical synthesis with control at sample level which entails potential uses in psychoacoustic experiments and synthesis with extreme precision (recap in Section~\ref{cap:conclusao}).
	\item Provide a powerful theoretical framework which can be used to synthesize musical pieces and albums.
	\item Provide approachability to the developed framework~\footnote{All
		the analytic relations presented in this article are implemented as public domain scripts,
		i.e. small computer programs using accessible technologies for distribution and validation.
		This constitutes the \massa\ toolbox, available in an open Git repository~\cite{gitBook}.
		These scripts are written in Python and make use of Numpy,
		which performs numerical routines efficiently (e.g. through LAPACK).
		Part of the scripts have been transcribed to JavaScript and native Python,
		which favors their use in Web browsers such as Firefox and
		Chromium~\cite{numpy, audiolab, tutpython, python}.
		These are all open technologies, published using licenses that allow for copy,
		distribution and use for research and generating derivatives.
		Hence, the work presented here aims at being compliant with recommended
		practices for availability and validation
		and should ease co-authorship processes~\cite{Raymond,Lessig}.}.
	\item Provide a didactic presentation of the content, which is highly multidisciplinary,
		involving signal processing, music, psycho-acoustics and programming.
\end{enumerate}

@@

\subsection{Synonymy, polysemy and theoretical frames (disclaimer)}
Given that the main topic of this article (the expression of musical elements
in PCM audio) is multidisciplinary and involves art,
the reader should be aware that many of the vocabulary admit different choices of terms and definitions.
More specifically, it is often often the case where many words can express the same concept
and where one word can carry different meanings.
This is a very deep issue which might receive a dedicated manuscript.
The reader might need to read the rest of this document to understand this
small selection of synonymy and polysemy in the literature,
but it is important to illustrate the point before the more dense sections:
\begin{itemize}
	\item A ``note'' can mean a pitch or an abstract construct with pitch and duration or a sound emitted from a musical instrument or a specific note in a score or a music.
	\item The sampling rate (discussed above) is also called the sampling frequency.
	\item An harmonic in a sound is most often a sinusoidal component which is is the harmonic series of the fundamental frequency. Many times, however, the terms harmonic and component are not distinguished. A harmonic can also be a note performed in a instrument by preventing certain overtones (components).
	\item Harmony can refer to chords or to note sets related to chords or even to ``harmony'' in a more general sense, as a kind of balance and consistency.
	\item A ``tremolo'' can mean different things, e.g. in a piano score, a tremolo is a fast alternation of two notes (pitches) while in computer music theory it is (most often) an oscillation of intensity.
\end{itemize}

We strived to avoid nomenclature clashes and the use of more terms than needed.
Also, there are many theoretical standpoints for the understanding of musical phenomena,
which is an evidence that most often there is not a single way to express or characterize musical structures.
Therefore, in this article, adjectives such as "often" "commonly" "frequently" are abundant and they would probably be even more numerous if we wanted to be pedantically precise.
Some of these issues are exposed when the context is convenient, such as in the first considerations of timbre.

@@

\subsection{Duration}
The sample frequency $f_s$ is defined as the number of samples in each second of the discrete-time signal. Let $T_i=\{t_i\}$ be an ordered set of real samples separated by $\delta_s=1/f_s$ seconds.
A musical note of duration $\Delta$ seconds can be expressed as a sequence $T_i^{\Delta}$ of $\lfloor \Delta . f_s \rfloor$ samples.
That is, the integer part of the multiplication is considered, and an error of at most $\delta_s$ missing seconds is admitted, which is usually fine for musical purposes. As an example, if $f_s=44.1kHz \;\;\Rightarrow\;\;\delta_s=\frac{1}{44100}\approx 23$ microseconds. Thus:

@@

\noindent where $\lambda_f=\frac{f_s}{f}=\frac{\delta_f}{\delta_s}$ is the number of samples in the period.

@@

Let $e_i$ be integers indexed by $i$ such that
$f=\varepsilon^{e_i} f_0$,
where $f_0$ is any fixed frequency.
The symmetric scales mentioned above can be expressed as:
\begin{equation}\label{escSim}
\begin{aligned}
	\text{chromatic}    & = E_i^c    & = \{e_i^c\}_0^{11}   & =  \{0,1,2,3,4,5,6,7,8,9,10,11\} & = \{i\}_0^{11}\\
	\text{whole tones}  & = E_i^{wt} & = \{e_i^{wt}\}_0^{5} & = \{0,2,4,6,8,10\}               & = \{2.i\}_0^{5} \\
	\text{minor thirds} & = E_i^{mt} & = \{e_i^{mt}\}_0^{3} & = \{0,3,6,9\}                    & = \{3.i\}_0^3 \\
	\text{major thirds} & = E_i^{Mt} & = \{e_i^{Mt}\}_0^{2} & = \{0,4,8\}                      & = \{4.i\}_0^2\\
	\text{tritones}     & = E_i^{tt} & = \{e_i^{tt}\}_0^{1} & = \{ 0, 6 \}                     & = \{6.i\}_0^1
\end{aligned}
\end{equation}

@@

\begin{equation}\label{eq:escalas}
\begin{aligned}
	\text{aeolian}    & = \text{natural minor scale}  & = & \\
			  & = E_i^m = \{e_i^m\}_0^6       & = & \{0,2,3,5,7,8,10\} \\
	\text{locrian}    & = E_i^{lo} = \{e_i^{lo}\}_0^6 & = & \{0,1,3,5,6,8,10\} \\ 
	\text{ionian}     & = \text{major scale}          & = &  \\
			  & = E_i^M = \{e_i^M\}_0^6       & = & \{0,2,4,5,7,9,11\} \\
	\text{dorian}     & = E_i^{d} = \{e_i^{d}\}_0^6   & = & \{0,2,3,5,7,9,10\} \\
	\text{phrygian}   & = E_i^{p} = \{e_i^{p}\}_0^6   & = & \{0,1,3,5,7,8,10\} \\
	\text{lydian}     & = E_i^{l}=\{e_i^{l}\}_0^6     & = & \{0,2,4,6,7,9,11\} \\
	\text{mixolydian} & = E_i^{mi} = \{e_i^{mi}\}_0^6 & = & \{0,2,4,5,7,9,10\}
\end{aligned}
\end{equation}

@@

Although it is not a traditional scale, the harmonic series is often used as such:
\begin{equation}\label{eq:serieHarmonica}
\begin{split}
H_i = & \{h_i\}_0^{19}= \\
    =  & \{ 0,12,19+0.02,  24,28-0.14, 31+0.2, 34-0.31, \\
                     & 36, 38+0.04,40-0.14, 42-0.49, 43+0.02, \\
                     & 44+0.41, 46-0.31, 47-0.12, \\
                     & 48, 49+0.05, 50+0.04, 51-0.02, 52-0.14   \}
\end{split}
\end{equation}

@@

\begin{equation}\label{eq:groups}
\begin{split}
\forall \;\; p_1,p_2 \in G \Rightarrow  \quad   & p_1 \bullet p_2  = p_3 \in G \\ 
     & \text{(closure property)} \\
\forall \;\; p_1,p_2,p_3 \in G \Rightarrow \quad & (p_1\bullet p_2)\bullet p_3  = p_1\bullet (p_2\bullet p_3) \\
     & \text{(associativity property)} \\
\exists \;\; e \in G :                  \quad    & p \bullet e  = e \bullet p \;,\;\;\; \forall\; p \in G  \\ 
     &  \text{(existence of the identity element)} \\
\forall \;\; p \in G, \;\exists\; p^{-1} :\quad  &  p\bullet p^{-1}i =p^{-1}\bullet p = e \\
     &  \text{(existence of the inverse element)}
\end{split}
\end{equation}

@@

The possibilities provided by this exposition pour from both the organization of knowledge and the ability to achieve sounds which are extremely true to the models.
For example, one can produce noises with arbitrary resolution of the spectrum and a musical note can be synthesized with the parameters (e.g. of a vibrato) updated sample-by-sample.
Furthermore, software for synthesis and processing of sounds for musical purposes
by standard restrict the bit depth to 16 or 24.
This is achievable in this presented framework but by standard Python uses more bits per floating point number.
These ``higher fidelity'' characteristics can be crucial
e.g. for psychoacoustic experiments or for the development of
interfaces to generate high quality musical sounds or pieces.
Simply put, it is compelling
for many scientific and artistic purposes.
The didactic potential of the framework is evident when noticed that:
\begin{itemize}
	\item the integrals and derivatives, ubiquitous in continuous signal processing,
are all replaced by summations in discrete signals,
which are more intuitive and does not require fluency in calculus.
	\item The equations and concepts are implemented as software which can be easily assembled and inspected.
\end{itemize}
\noindent In fact, this framework was used in a number of contexts, including courses, software implementations and making music~\cite{vimeoLM,vivace,repoDissertacao,dissertacao}.
